A Bayesian approach to burial networks to infer social changes in northeastern Taiwan during the European colonization period

### 2021-02-03 analysis/paper/paper.Rmd
edit based on PL's comments

<!--burial study- network analysis-->
Burial analysis by archaeologists is an approach to understand past social structures through the study of the physical traces of mortuary practices. The material cultures and biological records of burial behaviors can represent the social ranking or identities of the deceased [@Binford1971; @Drennan2010; @Saxe1970]. Despite criticism that manipulation of burial rituals by the living can cause a disconnect between a person's status in life and their status represented by burial contexts [@Hodder1980; @Pearson1982], burials can still provide valuable information to infer past societies [@Chapman2003]. By studying burial forms, grave goods, and ritual behaviors that structure the material configuration of burials, we can infer social status and relations between members of a community that give an insight to social structures [@Byrd1995; @Seikel2011]. Based on network science and graph theory, the recent development of social network analysis provides many new tools to visualize and analyze relationships in archaeological data [@Borgatti2009; @Brughmans2013]. Network science is the study of relational data where a phenomenon is conceptualized as a network through two steps, abstraction and representation [@Brandes2013; @Collar2015]. A network can be viewed as a patterned aggregation that includes individual elements (i.e. individual burials), pair-wise relationships (the dyads, for example burials with similar types and amounts of grave goods), and an overall structure showing global patterns represented in the data. Burials with the same prestige goods could indicate some underlying social relations where individuals share similar access to trade, exchange, and gifting networks according to their status [eg. @Coward2013, pp. 252]. This paper will use a novel Bayesian approach on exponential random graph model for networks to investigate the formation of relationship between burials in northeastern Taiwan to explore the indirect impact of foreign presence on social organization.

@@

<!-- archaeological background -->
Archaeological sites in northeastern Taiwan show evidence of accumulation patterns of imported goods, such as ornaments and porcelains, in residential areas and the use as burial goods around the same time as European contacts in the 17th century [@Chen2007; @LWandBM2020ornament]. Imported goods were treated as prestige or wealth in those Indigenous societies that were commonly mentioned in Spanish historical records [@Borao2009; @LiandWu2006]. For example, the document written by a Spanish Priest described that the Spanish soldier used carnelian beads to exchange natural resources with local Indigenous people since their high values in Indigenous culture [@LiandWu2006]. In addition, an Indigenous person possessing more imported goods may have been recognized as more influential or having higher status by their community [@LiandWu2006]. Despite we observed the uneven distribution of imported goods in burials, the degree of differentiation across burials over time remain unclear. The use of imported items as prestige goods in local Indigenous communities might occur earlier before a European presence became established, but was maintained and amplified after European arrival due to local competition for power [@Borao2009; @LiandWu2006; @LWandBM2020ornament; @Ueda2016]. A general model to summarize this situation in northeastern Taiwan might be that the influence of a colonial power, combined with high local values of imported goods, might lead to increased social inequality due to competition among individuals [@Brumfiel1994; @Clark1994]. The observed uneven distribution in Northeastern Taiwan may be explained as a result of unequal access to trade networks when Indigenous societies were involved in the complex trade network stimulated by the Europeans.

<!-- archaeological specific case study -->
In this study, we explore social changes in a pericolonial context at Kiwulan (Figure \@ref(fig:figure-KWL-map)), an Iron Age site in northeastern Taiwan, which covers the time before the European arrival, the presence of the Europeans, including the Spanish and the Dutch, in the 17th century, and finally the Chinese in the 19th century [@Chen2007]. We assume that social changes at Kiwulan would be supported by chronological differences in the structure of burial networks. We examine burial networks before and after the foreign contacts, and test our hypothesis that increased social inequality can be observed in the network after the European arrival. We ask: (1) did European colonial activities in 17th century Taiwan result in an increased social inequality in an Indigenous society in ways that can be detected by analysis of burial networks? (2) what are the major variables affecting or forming the higher degree of popularity (e.g. a few nodes have more relationships with other nodes) in the European contact period that might hint at social heterogeneity during this period? By answering these questions, this study helps to expand the use of statistical modeling for burials in understanding the indirect effects of a colonial presence on Indigenous groups. This is important to understand reactions to indirect impacts of European colonization.

@@

<!-- ERGM in archaeology, and limitations --> 
Exponential random graph models first appeared in archaeology with @Brughmans2014, who studied Iron age settlement patterns in Southern Spain by modeling inter-settlement visibility networks and visual control at 159 sites. Although their models do not fit perfectly with the observed network of archaeological data, they proposed that ERGM is a promising method for exploration and hypothesis testing for social processes. Similarly, @Amati2019 modeled three networks consisting of 15 sites (AD 100 to 400) in the Caribbean to explore interaction mechanisms, including proximity, inter-cultural items, and pottery types. By comparing hypothesized networks with the observed sites, they found that the presence of hub sites can be efficiently explained by multiple interdependent mechanisms instead of only one variable exclusively. However, those studies also point out some limitations of ERGM, such as sensitivity to missing data and less ability to handle uncertainty. In addition, it is difficult in ERGM to estimate model parameters and interpret the result due to intractable likelihood normalizing constants and model degeneracy [@Caimo2014; @Jin2013]. A normalizing constant is a function of the model parameter for making probability distributions integrate to one, which becomes harder to compute with larger set of networks [@Caimo2017bayesian]. This is also termed "doubly intractable" since both the likelihood normalizing constant and the marginal likelihood (the evidence of the posterior) are hard to derive [@Caimo2013; @Lyne2015]. Model degeneracy is another issue that probability models tend to overestimate a small number of extreme graphs by assigning too much weight, such as empty (all nodes unconnected) or complete graphs (all nodes connected) [@Schweinberger2011; @Caimo2014]. One solution to these limitations is available by implementing ERGMs in a Bayesian framework.

@@

<!-- Bayesian inference for Exponential Random Graph Models --> 
Bayesian approaches to Exponential Random Graph Models (ERGMs) are effective tools for network modeling by incorporating prior information about the network configurations to better understand dependencies of network variables and improve computational issues in ERGMs [@Caimo2017; @Lehmann2020]. Prior information is derived from previous data or assumptions of our data. The advantage that Bayesian modeling has over traditional ERGMs is the application of Markov chain Monte Carlo (MCMC) simulation using the approximate exchange algorithm [@Caimo2011]. Bayesian ERGMs avoid doubly-intractable computations by directly sampling from the not normalized part of the posterior, which alleviates the computational problems and gives a better convergence results. This enable us to deal with complicated dependence patterns with ease, providing better estimations for complex social network models with heterogeneous data [@Caimo2017; @Snijders2006]. After fitting an ERMG with an MCMC algorithm, the Bayesian approach generates posterior probabilities incorporating our sample data and prior information that estimates the effect of each ERGM parameter in our models [@Caimo2015; @Nemmers2019]. This allows uncertainty quantification by looking at the posterior mean and 95% credible intervals, or the odds ratios (conditional probability of observing an edge) that can replace p-values for assessing a null hypothesis [@Caimo2017bayesian]. The typical criteria for interpreting an significant effect of a parameter is that its credible interval does not contain zero, or an odds ratio is greater/less than one, meaning a significant positive/negative effect [@Harris2013, pp. 51]. In addition, Bayesian approaches are also useful to deal with missing data, which is often a problem leading to misinterpretation of networks, especially for archaeological studies. @Koskinen2010 shows that the effect of missing data can be reduced with Bayesian modeling that can predict, on the average, 80% of the ties when a third of data missing.

@@

<!-- archaeological setting -->
We analyzed burial data collected from the published excavation reports, and the original fieldwork notes for the upper component of Kiwulan, an Iron Age settlement (1350-1850 AD) [@Chen2007]. Excavations revealed abundant trade goods, including imported ceramics and stonewares, metal artifacts, glass beads, and stone beads, indicative of frequent trade activities [@Chen2007]. A total of 90 burials were unearthed that are mostly located in the middle section of the excavation area, which is the largest open area at Kiwulan that provides continuous stratigraphic sections suitable for temporal comparison (Figure \@ref(fig:figure-burial-map)). Burials are oriented in an east-west direction on the north side of the residential area, indicated by post-holes and _in-situ_ wooden posts, suggesting a well organized spatial arrangement. Previous studies report an uneven distribution of prestige goods across burials without agreement about whether this uneven distribution hints at vertical social differences. For example, @Cheng2008 interpreted the unequal distribution of glass beads, especially the gold-foil beads between burials, as evidence for hierarchy, indicating a stratified society. However, @Hsieh2012 suggested a relatively egalitarian structure based on comparative analysis of the frequencies of all burial goods. She found that the burials with high value goods were usually associated with elders, which might indicate achieved, rather than inherited, status. One important limitation of these previous studies is that they did not use analytical units suitable for comparing pre-European social organization with post-European social organization. Here, we adopt a new chronological framework for the burials to test if network configurations differ from the pre-European period to the post-European period. The discord over the discussion of Kiwulan social organization could be associated with chronological differences that has not yet been well studied.

@@

<!-- burial data and the definition of tie formation  --> 
We used burial goods in common to construct ties for representing a flow of good between burials. Since trade beads are commonly found across burials with substantial differences in quantities, we described each burial as having one of four levels, high, upper-middle, lower-middle, and low, according to their distributions across all burials (details in Supplementary Online Materials). For less frequent prestige goods, including imported ceramics, gold foils, and fish-shaped ornaments, we linked two burials when they possess each type of goods (i.e. presence or absence). Node attributes here include osteological data, such as age and sex, and cultural data, such as ritual pottery, and burial value index. Ritual pottery was identified as locally made ceramics placed above graves that suggests funeral feasting based on historical records. The burial value index is an important attribute for economical inequality that we assigned the burials into three levels, high (>`r high_value`, top 10 percent), above average (`r ave_value`-`r high_value`), and below average (<`r ave_value`), as an index of wealth. The number is the sum of values of each type of prestige good, which is calculate by the total number of a prestige item from burial context over the number of prestige item in a burial. Since burials tend to have multiple prestige goods in common, the network ties are weighted instead of binary data (the value 1 represents a tie and the value 0 otherwise) [@Snijders2011]. For example, if two burials have both low quantity of glass beads and porcelain in common, the tie is given a value of 2. Our networks are non-directed, which means ties have no orientation forming the relationship between actors to indicate a mutual relationship. The networks constructed based on this principles show that the network after the European presence has higher node degrees in general (Figure \@ref(fig:figure-network-diagrams)).

@@

<!-- Network construction and specification -->
We quantify the relations among burials and test our hypothesis of social change using the R programming language [@Rlanguage2019] with the bergm package [@Caimo2014], which is built upon the statnet package for model specification and simulation procedures [@Hunter2008; @Morris2008; @Handcock2008]. Table \@ref(tab:table-ergm-terms) lists the network parameters we used for the model specification with the corresponding archaeological evidence. Every parameter in an ERGM has an associated algorithm for computing the probability of observing relations in terms of grave goods between two burials. Based on our hypothesis, we model a network with increased social inequality that is represented by endogenous network effects: low transitivity and high popularity. We include burial-specific attributes as covariate effects for homophily, such as age, sex, ritual activity, and the degree of wealth, to explore and test the tendency of tie forming. For example, age-homophily means people of the same age tend to have the same burial goods. We also include the physical distance between burials as an indicator of a kinship-based relation since the deceased from the same family tend to be buried nearby [@LiandWu2006]. Our model could reveal the emergence of social inequality via the presence of a few individuals as network centers, having more relations with others. This would be indicated by high popularity or degree values in the network statistics, with covariates to control the preferential tendency of formation of relationship. According to our hypothesis, the burial evidence from after the European arrival will show higher popularity compared with the burial evidence from before European contact.

```{r table-ergm-terms}
ergm_terms <- readxl::read_excel(here::here("analysis", "data", "raw_data", "ergm_term.xlsx")) 
knitr::kable(ergm_terms,
             caption = "Network parameters in ERGMs used for model specifications with assocaited interpeation for burial relations.")
```

@@

<!--posterior stats for pre-E-->
We examine the estimates from the posterior distributions to compare their differences in structure of the simulated networks (Table \@ref(tab:table-posterior-stats); Figure \@ref(fig:figure-posterior-distribution)). For nodal covariates in the pre-European model, ritual-homophily has a significant effect on the formation of relations between burials, while burial value-homophily has a negative effect. This demonstrates that burials with ritual pottery tend to form relations, but burials in the same value level tend not to. Despite positive mean value for some covariates, such as age-homophily, sex-homophily, they do not show a significant tendency duo to value of zero in the confidence intervals. Similarly, the dyadic covariate, physical distance, shows no significant effect, indicating that physical proximity between burials does not affect the formation of relations. For the endogenous network effects, transitivity (gwesp) present significant positive effect, while popularity (gwdegree) demonstrates negative effect. The high positive value for transitivity suggests a tendency of burials to be clustered in closed transitive structures, indicative of the presence of multiple corporate groups sharing burial goods in common. In contrast, the strong negative popularity shows there is a tendency toward decentralization that reflects most burials have similar number of ties without any prominent ones. This might imply that individuals have equal access to trade goods in terms of the flow of goods.

@@

library(flextable)
# add column labels
ft <- flextable(posterior_esti_two_phases) %>% 
  set_header_labels(mean_pre = "pre-Euorpean", median_pre = "pre-Euorpean", 
                    `2.5%_pre` = "pre-Euorpean", `97.5%_pre` = "pre-Euorpean",
                    mean_post = "post-Euorpean", median_post = "post-Euorpean", 
                    `2.5%_post` = "post-Euorpean", `97.5%_post` = "post-Euorpean") %>%
  merge_at(i = 1, j = 2:5, part = "header") %>% 
  merge_at(i = 1, j = 6:9, part = "header") %>% 
  add_header_row(values = c("", "mean", "median", "2.5%", "97.5%",
                            "mean", "median", "2.5%", "97.5%"), top = FALSE ) %>% 
  align(j = 2:9, align = "right", part = "header") %>% 
  align(i = 1, align = "center", part = "header") %>% 
  width(j = ~ parameter, width = 1.3) %>% 
  width(j = 2:9, width = 0.65) %>% 
  set_caption("Estimated posterior means, medians, and 95% credible intervals for each network parameter of two models.")
ft
```

```{r figure-posterior-distribution, fig.cap = "Posterior density estimates for the parameters associated with edges, gwesp, gwdegree, and distance-dyadcov by phases. The pre-European group presents remarkedly larger values for the gwesp, but lower vlaues for the edges and gwdegree parameters comapred to the post-Eruoepan group. The distance parameters overlap between two groups."}
knitr::include_graphics(here::here("analysis", "figures", "006-posterior-distribution.png"))
```

<!--posterior stats for for post-E and their comparison-->
For the network model after the arrival of Europeans, the nodal covariates of burial value-homophily show significant positive effects, while the ritual-homophily and sex-homophily have a negative effect. This indicates the burials in the same wealth level tend to form relations. On the contrary, the same sex burials and burial with ritual pottery in common tend not to form relations. There are no significant effects for age-homophily and the dyadic covariate, physical distance. Similar to the pre-European network, the endogenous network variable, transitivity (gwesp) demonstrates a significant positive effect but much lower than the effect of pre-European network. In contrast, the popularity (gwdegree) shows a significantly higher positive effect than the effect of pre-European network. This means there is a tendency toward centralization that reflects a limited number of burials have many more ties than others. This implies that a few central individuals might have better access to trade goods and developed personal networks through wealth accumulation. In general, the post-European network model has a smaller transitivity effect and a positive popularity effect than the pre-European network model. This may suggest a reduced tendency toward clustering but high tendency toward centralization after the European presence. In addtion, both posterior estimates present symmetric distributions, with the posterior means close to posterior medians (Table \@ref(tab:table-posterior-stats)).

@@

```{r figure-bootstrap-CI, fig.cap = "The 95% bootstrap confidence intervals for the observed networks, the top one with 0 nodes removal, and networks with different number of node removal. The number of bootstrap samples is 1000."}
knitr::include_graphics(here::here("analysis", "figures", "007-bootstrap-CI.png"))
```

@@

The vertex bootstrap procedure with a t-test on the difference of network populations also shows that there are significant differences in network density(p = $`r pval_den`$), transitivity (p = $`r pval_trans`$), and popularity (p = $`r pval_star`$). Moreover, we explored the sample size effect by removing nodes at certain percents (5-40%) for both networks and compared their confidence intervals. The results (Figure \@ref(fig:figure-bootstrap-CI)) demonstrate a consistent difference between the two networks within 30 % node removals that suggest they are robust to the different sizes. In general, those statistical methods support the presence of anthropologically meaningful differences in the pre-and post-European networks. The procedure of cross-validation ensures our results can provide robust comparison between two networks without sample size effects.

```{r figure-bgof-two-phases, fig.cap = "Goodness-of-fit diagnostics for the pre-European model (left) and the post-European model (right). Boxplots represent distributions calculated on 100 network graphs simulated from the estimated posterior distribution. Red lines represent distributions of observed networks, and gray lines show the 95% intervals."}
library(magick)
pre_bgof_img <- image_read(here::here("analysis/figures/003-pre-bgof.png"))
post_bgof_img <- image_read(here::here("analysis/figures/004-post-bgof.png"))
image_append(c(pre_bgof_img, post_bgof_img))
```

@@

<!--anthopological implication at Kiwulan and limitation-->
At Kiwulan, changes in interburial networks may result from differential access to exotic trade enabled by European contact, gradually leading to increased social differentiation. The image of European power could be embedded into prestige objects pre-existing in local Indigenous culture where their values were amplified with concepts of wealth or power, since their values are contextually mutable and entangled in historical processes [@Aswani2003; @Thomas2009]. This may lead to ambitious individuals to competing with each other for traditional goods and accumulating wealth. It also demonstrates the agency of local Indigenous societies to incorporate and manipulate trade items in a pericolonial context with weak colonial control, gradually resulting in social changes from a more corporate to a more network strategy. We recognize that the meaning of the burial goods remains speculative without rich ethnographic data to support our interpretations. In addition, we acknowledge the fragmentation issue in our archaeological data that limit network interpretations [eg. @Erik2015], such as missing age and sex attributes for some burials at Kiwulan.

>>>>

<!--burial study- network analysis-->
Burial analysis by archaeologists is an approach to understand past social structures through the study of the physical traces of mortuary practices. The material culture and biological records of burial behaviors can represent the social ranking or identities of the deceased [@Binford1971; @Drennan2010; @Saxe1970]. Despite criticism that manipulation of burial rituals by the living can cause a disconnect between a person's status in life and their status represented by burial contexts [@Hodder1980; @Pearson1982], burials can still provide valuable information to infer past societies [@Chapman2003]. By studying burial forms, grave goods, and ritual behaviors that structure the material configuration of burials, we can infer social status and relations between members of a community that give an insight to social structures [@Byrd1995; @Seikel2011]. Based on network science and graph theory, the recent development of social network analysis provides many new tools to visualize and analyze relationships in archaeological data [@Borgatti2009; @Brughmans2013]. Network science is the study of relational data where a phenomenon is conceptualized as a network through two steps, abstraction and representation [@Brandes2013; @Collar2015]. A network can be viewed as a patterned aggregation that includes individual elements (i.e. individual burials), pair-wise relationships (the dyads, for example burials with similar types and amounts of grave goods), and an overall structure showing global patterns represented in the data. Burials with the same prestige goods could indicate some underlying social relations where individuals share similar access to trade, exchange, and gifting networks according to their status [eg. @Coward2013, pp. 252]. This paper will use a novel Bayesian approach on exponential random graph models for networks to investigate the formation of relationships between burials in northeastern Taiwan to explore the indirect impact of foreign presence on social organization.

@@

<!-- archaeological background -->
Archaeological sites in northeastern Taiwan show the evidence of accumulation patterns of imported goods, such as ornaments and porcelains, in residential areas and burial contexts around the same time as European contacts in the 17th century [@Chen2007; @LWandBM2020ornament]. Imported goods were treated as prestige or wealth in those Indigenous societies that were commonly mentioned in Spanish historical records in the 17th century (1624-1642) [@Borao2009; @LiandWu2006]. For example, a document written by a Spanish priest described a Spanish soldier using carnelian beads to exchange for natural resources with local Indigenous people, because of their high value in Indigenous culture [@LiandWu2006]. In addition, Spanish visitors observed that an Indigenous person possessing more imported goods may have been recognized as more influential or having higher status by their community [@LiandWu2006]. Despite the fact that we recorded an uneven distribution of imported goods in burials, the degree of differentiation across burials over time remains unclear. The introduction of foreign prestige goods to local Indigenous communities might have occurred before European presence became established, but was maintained and amplified after European arrival due to local competition for power [@Borao2009; @LiandWu2006; @LWandBM2020ornament; @Ueda2016]. A general model to summarize this situation in northeastern Taiwan might be that the influence of a colonial power, combined with high local values of imported goods, might lead to increased social inequality due to competition among individuals [@Brumfiel1994; @Clark1994]. The observed uneven distribution in Northeastern Taiwan may be explained as a result of unequal access to trade networks when Indigenous societies became more involved in the complex trade network stimulated by Europeans.

<!-- archaeological specific case study -->
In this study, we explore social changes in a pericolonial context at Kiwulan (Figure \@ref(fig:figure-KWL-map)), an Iron Age site in northeastern Taiwan, which was occupied from 14th to 19th centuries. This occupation period includes the time before the European arrival, the presence of the Europeans, including the Spanish and the Dutch, in the 17th century, and finally the Chinese in the 19th century [@Chen2007]. We assume that social changes at Kiwulan would be supported by chronological differences in the structure of burial networks. We examine burial networks before and after the foreign contacts, and test our hypothesis that increased social inequality can be observed in the network after the European arrival. We ask: (1) did European colonial activities in 17th century Taiwan result in an increased social inequality in an Indigenous society in ways that can be detected by analysis of burial networks? (2) what are the major variables affecting or forming the higher degree of popularity (e.g. a few nodes have more relationships with other nodes) in the European contact period that might hint at social heterogeneity during this period? By answering these questions, this study helps to expand the use of statistical modeling for burials in understanding the indirect effects of a colonial presence on Indigenous groups. This is important to understand reactions to indirect impacts of European colonization.

@@

<!-- ERGM in archaeology, and limitations --> 
Exponential random graph models (ERGMs) first appeared in archaeology with @Brughmans2014, who studied Iron age settlement patterns in Southern Spain by modeling inter-settlement visibility networks and visual control at 159 sites. Although their models do not fit perfectly with the observed network of archaeological data, they proposed that ERGM is a promising method for exploration and hypothesis testing for social processes. Similarly, @Amati2019 modeled three networks consisting of 15 sites (AD 100 to 400) in the Caribbean to explore interaction mechanisms, including proximity, inter-cultural items, and pottery types. By comparing hypothesized networks with the observed sites, they found that the presence of hub sites can be efficiently explained by multiple interdependent mechanisms instead of only one variable exclusively. However, those studies also point out some limitations of ERGM, such as sensitivity to missing data and less ability to handle uncertainty. In addition, it is difficult in ERGM to estimate model parameters and interpret the result due to intractable likelihood normalizing constants and model degeneracy [@Caimo2014; @Jin2013]. A normalizing constant is a function of the model parameter for making probability distributions integrate to one, which becomes harder to compute with larger set of networks [@Caimo2017bayesian]. This is also termed "doubly intractable" since both the likelihood normalizing constant and the marginal likelihood (the evidence of the posterior) are hard to derive [@Caimo2013; @Lyne2015]. Model degeneracy is another issue that probability models tend to overestimate a small number of extreme graphs by assigning too much weight, such as empty (all nodes unconnected) or complete graphs (all nodes connected) [@Schweinberger2011; @Caimo2014]. One solution to these limitations is available by implementing ERGMs in a Bayesian framework.

@@

<!-- Bayesian inference for Exponential Random Graph Models --> 
Bayesian approaches to ERGMs are effective tools for network modeling by incorporating prior information about the network configurations to better understand dependencies of network variables and improve computational issues in ERGMs [@Caimo2017; @Lehmann2020]. Prior information is derived from previous data or assumptions of our data. The advantage that Bayesian modeling has over traditional ERGMs is the application of Markov chain Monte Carlo (MCMC) simulation using the approximate exchange algorithm [@Caimo2011]. Bayesian ERGMs avoid doubly-intractable computations by directly sampling from the not normalized part of the posterior, where the acceptance probability is tractable that give a better convergence result. This enables us to deal with complicated dependence patterns with ease, providing better estimations for complex social network models with heterogeneous data [@Caimo2017; @Snijders2006]. After fitting an ERMG with an MCMC algorithm, the Bayesian approach generates posterior probabilities incorporating our sample data and prior information that estimates the effect of each ERGM parameter in our models [@Caimo2015; @Nemmers2019]. This allows uncertainty quantification by looking at the posterior mean and 95% credible intervals, or the odds ratios (conditional probability of observing an edge) that can replace p-values for assessing a null hypothesis [@Caimo2017bayesian]. The typical criteria for interpreting an significant effect of a parameter is that its credible interval does not contain zero, or an odds ratio is greater/less than one, meaning a significant positive/negative effect [@Harris2013, pp. 51]. In addition, Bayesian approaches are also useful to deal with missing data, which is often a problem leading to misinterpretation of networks, especially for archaeological studies. @Koskinen2010 shows that the effect of missing data can be reduced with Bayesian modeling that can predict, on the average, 80% of the ties when a third of data is missing.

@@

<!-- archaeological setting -->
We analyzed burial data collected from the published excavation reports, and the original fieldwork notes for the upper component of Kiwulan, an Iron Age settlement (1350-1850 AD) [@Chen2007]. Excavations revealed abundant trade goods, including imported ceramics and stonewares, metal artifacts, glass beads, and stone beads, indicative of frequent trade activities [@Chen2007]. A total of 90 burials were unearthed that are mostly located in the middle section of the excavation area, which is the largest open area at Kiwulan that provides continuous stratigraphic sections suitable for temporal comparison (Figure \@ref(fig:figure-burial-map)). Burials are oriented in an east-west direction on the north side of the residential area, indicated by post-holes and _in-situ_ wooden posts, suggesting a well organized spatial arrangement. Previous studies report an uneven distribution of prestige goods across burials without agreement about whether this uneven distribution hints at vertical social differences. For example, @Cheng2008 interpreted the unequal distribution of glass beads, especially the gold-foil beads between burials, as evidence for hierarchy, indicating a stratified society. However, @Hsieh2012 suggested a relatively egalitarian structure based on comparative analysis of the frequencies of all burial goods. She found that the burials with high value goods were usually associated with elders, which might indicate achieved, rather than inherited, status. One important limitation of these previous studies is that they did not use analytical units suitable for comparing pre-European social organization with post-European social organization. Here, we adopt a new chronological framework for the burials to test if network configurations differ from the pre-European period to the post-European period. The discord over the discussion of Kiwulan social organization could be associated with chronological differences that have not yet been well studied.

@@

<!-- burial data and the definition of tie formation  --> 
We used burial goods in common to construct ties for representing a flow of goods between burials. Since trade beads are commonly found across burials with substantial differences in quantities, we described each burial as having one of four levels, high, upper-middle, lower-middle, and low, according to their distributions across all burials (details in Supplementary Online Materials). For less frequent prestige goods, including imported ceramics, gold foils, and fish-shaped ornaments, we linked two burials when they possess each type of goods (i.e. presence or absence). Node attributes here include osteological data, such as age and sex, and cultural data, such as ritual pottery, and burial value index. Ritual pottery was identified as locally made ceramics placed above graves that suggests funeral feasting based on historical records. The burial value index is an important attribute for economical inequality that we assigned the burials into three levels, high (>`r high_value`, top 10 percent), above average (`r ave_value`-`r high_value`), and below average (<`r ave_value`), as an index of wealth. The number is the sum of values of each type of prestige good, which is calculated by the total number of a prestige item from burial context over the number of prestige items in a burial. Since burials tend to have multiple prestige goods in common, the network ties are weighted instead of binary data (the value 1 represents a tie and the value 0 otherwise) [@Snijders2011]. For example, if two burials have both low quantity of glass beads and porcelain in common, the tie is given a value of 2. Our networks are non-directed, which means ties have no orientation forming the relationship between actors to indicate a mutual relationship. The networks constructed based on these principles show that the network after the European presence has higher node degrees in general (Figure \@ref(fig:figure-network-diagrams)).

@@

<!-- Network construction and specification -->
We quantify the relations among burials and test our hypothesis of social change using the R programming language [@Rlanguage2019] with the bergm package [@Caimo2014], which is built upon the statnet package for model specification and simulation procedures [@Hunter2008; @Morris2008; @Handcock2008]. Table \@ref(tab:table-ergm-terms) lists the network parameters we used for the model specification with the corresponding archaeological evidence. Every parameter in an ERGM has an associated algorithm for computing the probability of observing relations in terms of grave goods between two burials. Based on our hypothesis, we model a network with increased social inequality that is represented by endogenous network effects: low transitivity and high popularity. We include burial-specific attributes as covariate effects for homophily, such as age, sex, ritual activity, and the degree of wealth, to explore and test the tendency of tie forming. For example, age-homophily means people of the same age tend to have the same burial goods. We also include the physical distance between burials as an indicator of a kinship-based relation since the deceased from the same family tend to be buried nearby [@LiandWu2006]. Our model could reveal the emergence of social inequality via the presence of a few individuals as network centers, having more relations with others. This would be indicated by high popularity or degree values in the network statistics, with covariates to control the preferential tendency of formation of relationships. According to our hypothesis, the burial evidence from after the European arrival will show higher popularity compared with the burial evidence from before European contact.

```{r table-ergm-terms}
ergm_terms <- readxl::read_excel(here::here("analysis", "data", "raw_data", "ergm_term.xlsx")) 
knitr::kable(ergm_terms,
             caption = "Network parameters in ERGMs used for model specifications with associated interpretation for burial relations.")
```

@@

<!--posterior stats for pre-E-->
We examine the estimates from the posterior distributions to compare their differences in structure of the simulated networks (Table \@ref(tab:table-posterior-stats); Figure \@ref(fig:figure-posterior-distribution)). For nodal covariates in the pre-European model, ritual-homophily has a significant effect on the formation of relations between burials, while burial value-homophily has a negative effect. This demonstrates that burials with ritual pottery tend to form relations, but burials in the same value level tend not to. Despite positive mean value for some covariates, such as age-homophily, sex-homophily, they do not show a significant tendency duo to value of zero in the confidence intervals. Similarly, the dyadic covariate, physical distance, shows no significant effect, indicating that physical proximity between burials does not affect the formation of relations. For the endogenous network effects, transitivity (gwesp) presents significant positive effect, while popularity (gwdegree) demonstrates negative effect. The high positive value for transitivity suggests a tendency of burials to be clustered in closed transitive structures, indicative of the presence of multiple corporate groups sharing burial goods in common. In contrast, the strong negative popularity shows there is a tendency toward decentralization that reflects most burials have similar number of ties without any prominent ones. This might imply that individuals have equal access to trade goods in terms of the flow of goods.

@@

library(flextable)
# add column labels
ft <- flextable(posterior_esti_two_phases) %>% 
  set_header_labels(mean_pre = "pre-European", median_pre = "pre-European", 
                    `2.5%_pre` = "pre-European", `97.5%_pre` = "pre-European",
                    mean_post = "post-European", median_post = "post-European", 
                    `2.5%_post` = "post-European", `97.5%_post` = "post-European") %>%
  merge_at(i = 1, j = 2:5, part = "header") %>% 
  merge_at(i = 1, j = 6:9, part = "header") %>% 
  add_header_row(values = c("", "mean", "median", "2.5%", "97.5%",
                            "mean", "median", "2.5%", "97.5%"), top = FALSE ) %>% 
  align(j = 2:9, align = "right", part = "header") %>% 
  align(i = 1, align = "center", part = "header") %>% 
  width(j = ~ parameter, width = 1.3) %>% 
  width(j = 2:9, width = 0.65) %>% 
  set_caption("Estimated posterior means, medians, and 95% credible intervals for each network parameter of two models.")
ft
```

```{r figure-posterior-distribution, fig.cap = "Posterior density estimates for the parameters associated with edges, gwesp, gwdegree, and distance-dyadcov by phases. The pre-European group presents remarkable larger values for the gwesp, but lower values for the edges and gwdegree parameters compared to the post-European group. The distance parameters overlap between two groups."}
knitr::include_graphics(here::here("analysis", "figures", "006-posterior-distribution.png"))
```

<!--posterior stats for for post-E and their comparison-->
For the network model after the arrival of Europeans, the nodal covariates of burial value-homophily show significant positive effects, while the ritual-homophily and sex-homophily have a negative effect. This indicates the burials in the same wealth level tend to form relations. On the contrary, the same sex burials and burial with ritual pottery in common tend not to form relations. There are no significant effects for age-homophily and the dyadic covariate, physical distance. Similar to the pre-European network, the endogenous network variable, transitivity (gwesp) demonstrates a significant positive effect but much lower than the effect of pre-European network. In contrast, the popularity (gwdegree) shows a significantly higher positive effect than the effect of the pre-European network. This means there is a tendency toward centralization that reflects a limited number of burials having many more ties than others. This implies that the presence of a better access to trade goods and the behavior of wealth accumulation or display in burial events. In general, the post-European network model has a smaller transitivity effect and a positive popularity effect than the pre-European network model. This may suggest a reduced tendency toward clustering but high tendency toward centralization after the European presence. In addtion, both posterior estimates present symmetric distributions, with the posterior means close to posterior medians (Table \@ref(tab:table-posterior-stats)).

@@

```{r figure-bootstrap-CI, fig.cap = "The 95% bootstrap confidence intervals for the observed networks, the top one with 0 nodes removal, and networks with different numbers of node removal. The number of bootstrap samples is 1000."}
knitr::include_graphics(here::here("analysis", "figures", "007-bootstrap-CI.png"))
```

@@

The vertex bootstrap procedure with a t-test on the difference of network populations also shows that there are significant differences in network density(p = $`r pval_den`$), transitivity (p = $`r pval_trans`$), and popularity (p = $`r pval_star`$). Moreover, we explored the sample size effect by removing nodes at certain percent (5-40%) for both networks and compared their confidence intervals. The results (Figure \@ref(fig:figure-bootstrap-CI)) demonstrate a consistent difference between the two networks within 30 % node removals that suggest they are robust to the different sizes. In general, those statistical methods support the presence of anthropologically meaningful differences in the pre-and post-European networks. The procedure of cross-validation ensures our results can provide robust comparison between two networks without sample size effects.

```{r figure-bgof-two-phases, fig.cap = "Goodness-of-fit diagnostics for the pre-European model (top) and the post-European model (below). Boxplots represent distributions calculated on 100 network graphs simulated from the estimated posterior distribution. Red lines represent distributions of observed networks, and gray lines show the 95% intervals."}
library(magick)
pre_bgof_img <- image_read(here::here("analysis/figures/003-pre-bgof.png"))
post_bgof_img <- image_read(here::here("analysis/figures/004-post-bgof.png"))
image_append(c(pre_bgof_img, post_bgof_img), stack = T)
```

@@

<!--anthopological implication at Kiwulan and limitation-->
At Kiwulan, changes in interburial networks may result from differential access to exotic trade enabled by European contact, gradually leading to increased social differentiation. The image of European power could be embedded into prestige objects pre-existing in local Indigenous culture where their values were amplified with concepts of wealth or power, since their values are contextually mutable and entangled in historical processes [@Aswani2003; @Thomas2009]. This may lead to ambitious individuals to competing with each other for traditional goods and accumulating wealth. It also demonstrates that local Indigenous societies could actively incorporate and manipulate trade items in a pericolonial context with weak colonial control, gradually resulting in social changes from a more corporate to a more network strategy. We recognize that the meaning of the burial goods remains speculative without rich ethnographic data to support our interpretations. Also, the fragmentation issue in our archaeological data could limit network interpretations [eg. @Erik2015], such as missing age and sex attributes for some burials at Kiwulan. Human skeletal remains can provide information about health for interpreting social status; however, the preservation of Kiwulan human remains was generally not good to yield useful data.
